---
title: "CPD ISR — Unit 189 (Narcotics) 2016 Pipeline for Shiny"
author: "Devin Rooney"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: united
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6, cache = FALSE
)
```

# Overview

End-to-end pipeline from raw ISR parquet to a single `.rds` file consumed
by the Shiny app. Runs start-to-finish in a fresh R session.

Pipeline steps:

1. Load and deduplicate ISR stops
2. Load CPD beat shapefile (used for ungeocoded stop salvage)
3. Impute reliable event IDs
4. Salvage ungeocoded stops via BEAT match
5. Filter to Unit 189 (Narcotics), 2016
6. Build officer co-event network (NODES, EDGE_CONTRIBS)
7. Build per-officer-per-stop scope table (SCOPE_FOR_SHINY)
8. Save consolidated `.rds`

---

# Section 1: Setup

```{r setup-packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(arrow)
  library(sf)
  library(igraph)
})

options(tigris_use_cache = TRUE, scipen = 999)
set.seed(1848)

# --- Paths ---
DIR_DATA   <- "../data"
DIR_SHINY  <- "shiny_app"
dir.create(DIR_SHINY, showWarnings = FALSE, recursive = TRUE)

BEATS_PATH <- file.path(DIR_DATA, "CPD_beats.geojson")

# --- Constants ---
CRS_PROJ <- 26971       # NAD83 / Illinois East (ftUS)

# Event imputation parameters
WINDOW_SEC <- 30 * 60   # 30 minutes
RADIUS_FT  <- 250       # 250 feet

# Scope parameters
UNIT_ID            <- "189"                                 # Narcotics
SCOPE_START        <- as.POSIXct("2016-01-01", tz = "UTC")
SCOPE_END          <- as.POSIXct("2017-01-01", tz = "UTC")  # exclusive
EDGE_DURATION_DAYS <- 14
EDGE_DURATION_SEC  <- EDGE_DURATION_DAYS * 86400
CORE_MIN           <- 10                                    # tier classification
```

---

# Section 2: Load ISR

```{r s01-load-isr}
message("Stage 1: loading and deduplicating ISR parquet...")

ISR <- read_parquet(file.path(DIR_DATA, "ISR_2016_2023.parquet"))

ISR <- ISR |>
  mutate(
    DISTRICT = str_pad(as.character(DISTRICT), 2, "left", "0"),
    RACE = case_when(
      RACE_CODE_CD == "WHI" ~ "WHT",
      TRUE ~ RACE_CODE_CD
    ),
    YEAR       = as.integer(format(CONTACT_DATE, "%Y")),
    HIT        = as.integer(SEARCH_CONTRABAND_FOUND_I == "Y"),
    SEARCHED   = as.integer(SEARCH_I == "Y"),
    OFFICER_ID = as.character(FO_EMPLOYEE_ID)
  )

# Keep only the latest row per CARD_NO (some have revisions)
ISR <- ISR |>
  group_by(CARD_NO) |>
  slice_tail(n = 1) |>
  ungroup()

ISR <- ISR |>
  mutate(
    BEAT_PAD     = str_pad(as.character(BEAT),     4, "left", "0"),
    RES_BEAT_PAD = str_pad(as.character(RES_BEAT), 4, "left", "0")
  )

# Count how many stops share each EVENT_NO (for clustering decisions)
EVENT_COUNTS <- ISR |>
  filter(!is.na(EVENT_NO), EVENT_NO != "") |>
  count(EVENT_NO, name = "N_STOPS_IN_EVENT")

ISR <- ISR |>
  left_join(EVENT_COUNTS, by = "EVENT_NO") |>
  mutate(N_STOPS_IN_EVENT = replace_na(N_STOPS_IN_EVENT, 1L))
```

```{r s02-load-beats}
message("Stage 2: loading beat shapefile...")

BEATS <- st_read(BEATS_PATH, quiet = TRUE) |>
  st_transform(CRS_PROJ) |>
  mutate(BEAT = str_pad(as.character(beat_num), 4, "left", "0")) |>
  filter(!substr(BEAT, 1, 2) %in% c("31", "41", "51")) |>
  select(BEAT, geometry) |>
  st_make_valid()

stopifnot(anyDuplicated(BEATS$BEAT) == 0)
```

---

# Section 3: Event ID imputation

The raw `EVENT_NO` is unreliable: short codes like "00059" and "00000"
are reused across unrelated stops. Cluster stops sharing the same
`EVENT_NO` that fall within 250 ft and 30 min of each other (single-
linkage). For stops with missing `EVENT_NO`, substitute
`SUPV_EMPLOYEE_ID` as the grouping key.

```{r s03-project-and-flag}
message("Stage 3: projecting geocoded stops and flagging missing IDs...")

cluster_event <- function(df, radius_ft, window_sec) {
  n <- nrow(df)
  if (n == 1) return(1L)
  dx <- outer(df$X, df$X, "-")
  dy <- outer(df$Y, df$Y, "-")
  dist_mat <- sqrt(dx^2 + dy^2)
  time_mat <- abs(outer(as.numeric(df$CONTACT_DATE),
                        as.numeric(df$CONTACT_DATE), "-"))
  adj <- (dist_mat <= radius_ft) & (time_mat <= window_sec)
  g <- graph_from_adjacency_matrix(adj, mode = "undirected", diag = FALSE)
  components(g)$membership
}

ISR_PROJ <- ISR |>
  filter(!is.na(lat), !is.na(long)) |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE) |>
  st_transform(CRS_PROJ)

coords <- st_coordinates(ISR_PROJ)
ISR_PROJ <- ISR_PROJ |>
  st_drop_geometry() |>
  mutate(X = coords[, 1], Y = coords[, 2])

ISR_PROJ <- ISR_PROJ |>
  mutate(EVENT_MISSING = is.na(EVENT_NO) | EVENT_NO == "" | EVENT_NO == "00000")

message("  Real EVENT_NO stops: ", sum(!ISR_PROJ$EVENT_MISSING))
message("  Missing-ID stops:    ", sum(ISR_PROJ$EVENT_MISSING))
```

```{r s04-cluster-real}
message("Stage 4: clustering real EVENT_NO stops...")

REAL_MULTI <- ISR_PROJ |>
  filter(!EVENT_MISSING, N_STOPS_IN_EVENT > 1) |>
  arrange(EVENT_NO, CONTACT_DATE) |>
  group_by(EVENT_NO) |>
  mutate(CLUSTER = cluster_event(pick(everything()),
                                 RADIUS_FT, WINDOW_SEC)) |>
  ungroup()

REAL_MULTI <- REAL_MULTI |>
  group_by(EVENT_NO) |>
  mutate(N_CLUSTERS = n_distinct(CLUSTER)) |>
  group_by(EVENT_NO, CLUSTER) |>
  mutate(CLUSTER_FIRST_TIME = min(CONTACT_DATE)) |>
  group_by(EVENT_NO) |>
  mutate(CLUSTER_RANK = dense_rank(CLUSTER_FIRST_TIME),
         SUFFIX = if_else(N_CLUSTERS == 1, "_0",
                          paste0("_", CLUSTER_RANK)),
         IMPUTED_EVENT_ID = paste0(EVENT_NO, SUFFIX)) |>
  ungroup() |>
  select(CARD_NO, IMPUTED_EVENT_ID)

REAL_SINGLETON <- ISR_PROJ |>
  filter(!EVENT_MISSING, N_STOPS_IN_EVENT == 1) |>
  mutate(IMPUTED_EVENT_ID = paste0(EVENT_NO, "_0")) |>
  select(CARD_NO, IMPUTED_EVENT_ID)
```

```{r s05-cluster-missing}
message("Stage 5: clustering missing-ID stops via SUPV_EMPLOYEE_ID...")

MISSING_NO_SUPV <- ISR_PROJ |>
  filter(EVENT_MISSING, is.na(SUPV_EMPLOYEE_ID) | SUPV_EMPLOYEE_ID == "") |>
  mutate(IMPUTED_EVENT_ID = paste0("NOEVENT_NOSUPV_", CARD_NO)) |>
  select(CARD_NO, IMPUTED_EVENT_ID)

MISSING_WITH_SUPV <- ISR_PROJ |>
  filter(EVENT_MISSING,
         !is.na(SUPV_EMPLOYEE_ID), SUPV_EMPLOYEE_ID != "") |>
  arrange(SUPV_EMPLOYEE_ID, CONTACT_DATE) |>
  group_by(SUPV_EMPLOYEE_ID) |>
  mutate(CLUSTER = cluster_event(pick(everything()),
                                 RADIUS_FT, WINDOW_SEC)) |>
  ungroup()

MISSING_WITH_SUPV <- MISSING_WITH_SUPV |>
  group_by(SUPV_EMPLOYEE_ID, CLUSTER) |>
  mutate(CLUSTER_FIRST_TIME = min(CONTACT_DATE),
         CLUSTER_SIZE = n()) |>
  group_by(SUPV_EMPLOYEE_ID) |>
  mutate(CLUSTER_RANK = dense_rank(CLUSTER_FIRST_TIME)) |>
  ungroup() |>
  mutate(IMPUTED_EVENT_ID = if_else(
    CLUSTER_SIZE == 1,
    paste0("NOEVENT_", SUPV_EMPLOYEE_ID, "_S", CARD_NO),
    paste0("NOEVENT_", SUPV_EMPLOYEE_ID, "_", CLUSTER_RANK)
  )) |>
  select(CARD_NO, IMPUTED_EVENT_ID)
```

```{r s06-combine}
message("Stage 6: combining and attaching IMPUTED_EVENT_ID to ISR...")

IMPUTED <- bind_rows(
  REAL_MULTI,
  REAL_SINGLETON,
  MISSING_NO_SUPV,
  MISSING_WITH_SUPV
)

NO_GEO <- ISR |>
  anti_join(IMPUTED, by = "CARD_NO") |>
  mutate(IMPUTED_EVENT_ID = paste0("NOGEO_", CARD_NO)) |>
  select(CARD_NO, IMPUTED_EVENT_ID)

IMPUTED <- bind_rows(IMPUTED, NO_GEO)

stopifnot(nrow(IMPUTED) == nrow(ISR))
stopifnot(anyDuplicated(IMPUTED$CARD_NO) == 0)

ISR <- ISR |>
  left_join(IMPUTED, by = "CARD_NO")

message("  IMPUTED_EVENT_ID attached. Unique imputed events: ",
        n_distinct(ISR$IMPUTED_EVENT_ID))
```

---

# Section 4: Salvage ungeocoded stops via BEAT match

Many ISR stops lack lat/long but have a BEAT and EVENT_NO. If a
geocoded cluster shares the same EVENT_NO and BEAT and is within the
30-min window, attach the ungeocoded stop to that cluster.

```{r s07-find-matches}
message("Stage 7: salvaging ungeocoded stops via BEAT match...")

UNGEO_CANDIDATES <- ISR |>
  filter((is.na(lat) | is.na(long)),
         !is.na(EVENT_NO), EVENT_NO != "", EVENT_NO != "00000",
         !is.na(BEAT_PAD), BEAT_PAD != "") |>
  select(CARD_NO, EVENT_NO, CONTACT_DATE, BEAT_PAD,
         IMPUTED_EVENT_ID_OLD = IMPUTED_EVENT_ID)

GEOCODED_CLUSTERS <- ISR |>
  filter(!is.na(lat), !is.na(long),
         !str_starts(IMPUTED_EVENT_ID, "NOEVENT_"),
         !str_starts(IMPUTED_EVENT_ID, "NOGEO_")) |>
  select(EVENT_NO, IMPUTED_EVENT_ID, CONTACT_DATE, BEAT_PAD)

MATCHES <- UNGEO_CANDIDATES |>
  inner_join(GEOCODED_CLUSTERS,
             by = c("EVENT_NO", "BEAT_PAD"),
             suffix = c("_UG", "_GC"),
             relationship = "many-to-many") |>
  mutate(TIME_DIFF = abs(as.numeric(CONTACT_DATE_UG) -
                         as.numeric(CONTACT_DATE_GC))) |>
  filter(TIME_DIFF <= WINDOW_SEC)

MATCHES_BY_CLUSTER <- MATCHES |>
  group_by(CARD_NO, IMPUTED_EVENT_ID) |>
  summarise(MIN_TIME_DIFF = min(TIME_DIFF), .groups = "drop")

SALVAGE <- MATCHES_BY_CLUSTER |>
  group_by(CARD_NO) |>
  arrange(MIN_TIME_DIFF, IMPUTED_EVENT_ID) |>
  slice_head(n = 1) |>
  ungroup() |>
  select(CARD_NO,
         IMPUTED_EVENT_ID_NEW = IMPUTED_EVENT_ID)

ISR <- ISR |>
  left_join(SALVAGE, by = "CARD_NO") |>
  mutate(
    IMPUTED_EVENT_ID = if_else(!is.na(IMPUTED_EVENT_ID_NEW),
                                IMPUTED_EVENT_ID_NEW,
                                IMPUTED_EVENT_ID)
  ) |>
  select(-IMPUTED_EVENT_ID_NEW)

message("  Final unique imputed events: ",
        n_distinct(ISR$IMPUTED_EVENT_ID))
```

---

# Section 5: Filter to scope

```{r s09-scope}
message("Stage 9: filtering to Unit ", UNIT_ID, ", ", format(SCOPE_START),
        " to ", format(SCOPE_END), "...")

SCOPE <- ISR |>
  filter(CPD_UNIT_NO == UNIT_ID,
         CONTACT_DATE >= SCOPE_START,
         CONTACT_DATE <  SCOPE_END)

message("  Stops in scope: ", nrow(SCOPE))
```

---

# Section 6: Build officer co-event network

Pivot to long form (one row per officer-stop), then build node and edge
tables. Each stop is attributed to BOTH the FO and SO if both are
present.

```{r s10-build-network}
SCOPE_LONG <- SCOPE |>
  select(CARD_NO, IMPUTED_EVENT_ID, CONTACT_DATE,
         FO_EMPLOYEE_ID, SO_EMPLOYEE_ID,
         SEARCHED, HIT, RACE) |>
  pivot_longer(c(FO_EMPLOYEE_ID, SO_EMPLOYEE_ID),
               names_to = "ROLE", values_to = "OFFICER_ID") |>
  filter(!is.na(OFFICER_ID), OFFICER_ID != "")

NODES <- SCOPE_LONG |>
  group_by(OFFICER_ID) |>
  summarise(N_STOPS    = n(),
            FIRST_SEEN = min(CONTACT_DATE),
            LAST_SEEN  = max(CONTACT_DATE),
            .groups    = "drop") |>
  mutate(TIER = if_else(N_STOPS >= CORE_MIN, "core", "bridge"))

message("  Officers: ", nrow(NODES),
        " (core: ", sum(NODES$TIER == "core"),
        ", bridge: ", sum(NODES$TIER == "bridge"), ")")

# Multi-officer events drive edges
EVENTS <- SCOPE_LONG |>
  group_by(IMPUTED_EVENT_ID) |>
  summarise(EVENT_TIME = min(CONTACT_DATE),
            N_OFFICERS = n_distinct(OFFICER_ID),
            .groups    = "drop") |>
  filter(N_OFFICERS >= 2)

# Each multi-officer event yields one row per officer pair (alphabetical)
EDGE_CONTRIBS <- SCOPE_LONG |>
  semi_join(EVENTS, by = "IMPUTED_EVENT_ID") |>
  select(IMPUTED_EVENT_ID, OFFICER_ID) |>
  distinct() |>
  inner_join(
    SCOPE_LONG |> semi_join(EVENTS, by = "IMPUTED_EVENT_ID") |>
      select(IMPUTED_EVENT_ID, OFFICER_ID) |> distinct(),
    by = "IMPUTED_EVENT_ID",
    suffix = c("_FROM", "_TO"),
    relationship = "many-to-many"
  ) |>
  filter(OFFICER_ID_FROM < OFFICER_ID_TO) |>
  rename(FROM = OFFICER_ID_FROM, TO = OFFICER_ID_TO) |>
  left_join(EVENTS |> select(IMPUTED_EVENT_ID, EVENT_TIME),
            by = "IMPUTED_EVENT_ID")

message("  Multi-officer events: ", nrow(EVENTS))
message("  Edge contributions:   ", nrow(EDGE_CONTRIBS))
```

---

# Section 7: Build per-officer-per-stop scope table

This is the table the Shiny app uses to compute per-officer hit rate,
search rate, racial breakdowns, and stop counts within any brushed
time window. One row per (officer, stop), so a stop with two officers
appears twice — once attributed to each officer.

```{r s11-scope-for-shiny}
SCOPE_FOR_SHINY <- SCOPE_LONG |>
  select(OFFICER_ID, CARD_NO, IMPUTED_EVENT_ID, CONTACT_DATE,
         SEARCHED, HIT, RACE) |>
  arrange(CONTACT_DATE)

message("  SCOPE_FOR_SHINY rows (officer-stop pairs): ", nrow(SCOPE_FOR_SHINY))
```

---

# Section 8: Save consolidated .rds

```{r s12-save}
saveRDS(
  list(
    NODES           = NODES,
    EDGE_CONTRIBS   = EDGE_CONTRIBS,
    SCOPE_FOR_SHINY = SCOPE_FOR_SHINY
  ),
  file = file.path(DIR_SHINY, "data.rds")
)

message("Wrote ", file.path(DIR_SHINY, "data.rds"))
```

```{r s13-summary}
list(
  unit               = UNIT_ID,
  window             = paste(format(SCOPE_START), "to", format(SCOPE_END)),
  stops              = nrow(SCOPE),
  officers           = nrow(NODES),
  officers_core      = sum(NODES$TIER == "core"),
  officers_bridge    = sum(NODES$TIER == "bridge"),
  events_multi       = nrow(EVENTS),
  edge_contribs      = nrow(EDGE_CONTRIBS),
  scope_for_shiny    = nrow(SCOPE_FOR_SHINY),
  edge_duration_days = EDGE_DURATION_DAYS
) |> str()
```
